---
title: "A Tiny Iroshizuku Ink Shop in R"
description: "Turning Pilot's beautifully named fountain pen inks into a tiny Japanese stationery-shop display with ggplot2."
author: "Chi"
date: "2026-08-15"
categories:
  - r
  - ggplot
  - dataviz
  - color
  - japan
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
editor: 
  markdown: 
    wrap: sentence
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(scales)
library(colorspace)
```

## 色彩雫

I just discovered that the R script I had written while back, which was a hand-curated colour palette of Pilot Iroshizuku fountain pen inks.

Pilot describes Iroshizuku as a combination of iro = colour, and shizuku = droplet.
The individual inks take their names from Japanese landscapes, plants, seasons, and other bits of nature.

The names don't simply tell you what colour an ink is.

- 月夜 isn't just dark blue. It is moonlit night.
- 花筏 isn't just pink. It evokes cherry blossom petals floating together on water like a little raft.
- 冬将軍 - literally the Winter General - somehow becomes a cool gray.

So naturally, I turned them into data!

## The data

This is a small hand-curated dataset of 24 Iroshizuku inks.
The hex colours are approximate representations rather than measurements of the physical inks.
Fountain pen ink is much more complicated than a single hex value: paper, nib width, saturation, shading, sheen, and lighting all change how an ink appears.

The ink names and descriptions are based on Pilot's [Iroshizuku collection](https://www.pilotpen.eu/our-universes/fine-writing/iroshizuku-inks/).
The hex values are approximate colours used for this visualization rather than official measured colour values.

But hex is enough for a little plotting experiment!

```{r}
#| label: iroshizuku-data

iroshizuku_colors <- tibble(
  ink_name = c(
    "Ajisai", "Asagao", "Konpeki", "Amairo", "Kujaku", "Rikka",
    "Tsukiyo", "Shinkai", "Syoro", "Shinryoku", "Suigyoku", "Takesumi",
    "Fuyusyogun", "Chikurin", "Hotarubi", "Hanaikada", "Murasakishikibu",
    "Yamabudo", "Momiji", "Fuyugaki", "Yuyake", "Toro", "Yamaguri", "Syungyo"
  ),
  ink_name_japanese = c(
    "紫陽花", "朝顔", "紺碧", "天色", "孔雀", "立夏",
    "月夜", "深海", "松露", "深緑", "翠玉", "竹炭",
    "冬将軍", "竹林", "蛍火", "花筏", "紫式部",
    "山葡萄", "紅葉", "冬柿", "夕焼け", "灯籠", "山栗", "春暁"
  ),
  hex = c(
    "#1255A2", "#04318E", "#0368B4", "#00A0DF", "#028986", "#1A7DA5",
    "#016D8C", "#1C3A65", "#077D5E", "#007E4F", "#037261", "#1E1D1E",
    "#6A869A", "#94BD4E", "#D9DA26", "#ED7E93", "#765FA8", "#660D5B",
    "#E12E2C", "#EA5A10", "#EF881F", "#F0B018", "#5B4532", "#674F4D"
  ),
  description = c(
    "Hydrangea",
    "Morning Glory",
    "Deep Cerulean Blue",
    "Sky Blue",
    "Peacock",
    "Early Summer",
    "Moonlit Night",
    "Deep Sea",
    "Dew on Pine Tree",
    "Forest Green",
    "Emerald",
    "Bamboo Charcoal",
    "Winter Commander",
    "Bamboo Forest",
    "Firefly Glow",
    "Floating Cherry Blossoms",
    "Murasaki Shikibu",
    "Wild Grape Vine",
    "Autumn Maple Leaves",
    "Winter Persimmon",
    "Sunset Glow",
    "Lantern Light",
    "Wild Chestnut",
    "Spring Dawn"
  )
)
```

## First: just the colours

Before making bottles, it is useful to see the palette itself.

```{r}
#| label: simple-palette
#| fig-width: 10
#| fig-height: 7
#| fig-cap: "The Iroshizuku palette in its original order."
#| echo: false

palette_plot <- iroshizuku_colors |>
  mutate(
    idx = row_number() - 1,
    col = idx %% 6,
    row = 3 - idx %/% 6
  )

ggplot(palette_plot) +
  geom_tile(
    aes(
      x = col,
      y = row,
      fill = hex
    ),
    width = 0.92,
    height = 0.82
  ) +
  geom_text(
    aes(
      x = col,
      y = row + 0.05,
      label = ink_name_japanese
    ),
    family = "Hiragino Sans",
    color = "white",
    fontface = "bold",
    size = 6
  ) +
  geom_text(
    aes(
      x = col,
      y = row - 0.17,
      label = ink_name
    ),
    family = "Baskerville",
    color = "white",
    size = 3
  ) +
  scale_fill_identity() +
  coord_equal() +
  theme_void()
```

This already makes a nice colour chart, but the ordering feels fairly arbitrary.
So I wanted to see what would happen if the inks were arranged by colour rather than by catalogue order.

## Sorting colours perceptually

RGB is useful for screens, but it isn't especially good at representing how humans perceive differences between colours.
For sorting this palette, I converted the hex values into **HCL colour space** using the `colorspace` package.

HCL separates colour into:

- **Hue** — roughly which colour family it belongs to
- **Chroma** — how colourful or saturated it feels
- **Luminance** — how light or dark it appears

That makes it much nicer for arranging colours in a visually coherent sequence.

```{r}
#| label: color-space
#| echo: true

hcl_coords <- coords(
  as(hex2RGB(iroshizuku_colors$hex), "polarLUV")
) |>
  as_tibble()

ink_plot <- iroshizuku_colors |>
  bind_cols(hcl_coords) |>
  mutate(
    # Rotate the hue wheel so the sequence starts near blue / teal.
    hue_sort = (H - 220) %% 360
  ) |>
  arrange(hue_sort, desc(L), desc(C)) |>
  mutate(
    plot_order = row_number() - 1,
    col = plot_order %% 6,
    row_raw = plot_order %/% 6,
    row = max(row_raw) - row_raw
  )
```

The exact order is not scientifically important.
I rotated the hue wheel so that the display begins around the blue-green part of the palette, simply because it produces a calmer visual flow for this particular set of inks.

## The tiny ink shop

I wanted the palette to look like a tiny Japanese stationery-shop display: rows of ink bottles, wooden shelves, paper labels, and little price cards.

The whole thing is still just `ggplot2`.

```{r}
#| label: ink-shop
#| fig-width: 11
#| fig-height: 8
#| fig-cap: "A tiny Iroshizuku ink shop, arranged in perceptual colour order."
#| echo: false
#| warning: false
#| message: false


shelves <- ink_plot |>
  distinct(row) |>
  mutate(shelf_y = row - 0.46)

ggplot() +
  
  # shop backing panel
  annotate(
    "rect",
    xmin = -0.90,
    xmax = 5.90,
    ymin = -0.82,
    ymax = 4.02,
    fill = "#F3EDE3",
    color = "#CCBEAC",
    linewidth = 0.7
  ) +
  
  # inner display wall
  annotate(
    "rect",
    xmin = -0.72,
    xmax = 5.72,
    ymin = -0.65,
    ymax = 3.72,
    fill = "#ECE4D7",
    color = "#D4C7B6",
    linewidth = 0.35
  ) +
  
  # wooden side posts
  annotate(
    "rect",
    xmin = -0.86,
    xmax = -0.70,
    ymin = -0.65,
    ymax = 3.72,
    fill = "#896448",
    color = NA
  ) +
  annotate(
    "rect",
    xmin = 5.70,
    xmax = 5.86,
    ymin = -0.65,
    ymax = 3.55,
    fill = "#896448",
    color = NA
  ) +
  
  # top rail
  annotate(
    "segment",
    x = -0.70,
    xend = 5.70,
    y = 3.54,
    yend = 3.54,
    color = "#B8A890",
    linewidth = 1
  ) +
  
  # shelf boards
  geom_rect(
    data = shelves,
    aes(
      xmin = -0.72,
      xmax = 5.72,
      ymin = shelf_y - 0.04,
      ymax = shelf_y + 0.04
    ),
    fill = "#9A714C",
    color = "#6F5038",
    linewidth = 0.3
  ) +
  
  # shelf front edge
  geom_rect(
    data = shelves,
    aes(
      xmin = -0.72,
      xmax = 5.72,
      ymin = shelf_y - 0.09,
      ymax = shelf_y - 0.04
    ),
    fill = alpha("#4B3525", 0.16),
    color = NA
  ) +
  
  # bottle shadow
  geom_rect(
    data = ink_plot,
    aes(
      xmin = col - 0.31,
      xmax = col + 0.31,
      ymin = row - 0.46,
      ymax = row - 0.42
    ),
    fill = alpha("black", 0.08),
    color = NA
  ) +
  
  # bottle body
  geom_rect(
    data = ink_plot,
    aes(
      xmin = col - 0.40,
      xmax = col + 0.40,
      ymin = row - 0.42,
      ymax = row + 0.08,
      fill = hex
    ),
    color = "#403A35",
    linewidth = 0.45
  ) +
  
  # shoulders
  geom_rect(
    data = ink_plot,
    aes(
      xmin = col - 0.28,
      xmax = col + 0.28,
      ymin = row + 0.08,
      ymax = row + 0.19,
      fill = hex
    ),
    color = "#403A35",
    linewidth = 0.40
  ) +
  
  # neck
  geom_rect(
    data = ink_plot,
    aes(
      xmin = col - 0.18,
      xmax = col + 0.18,
      ymin = row + 0.19,
      ymax = row + 0.29,
      fill = hex
    ),
    color = "#403A35",
    linewidth = 0.35
  ) +
  
  # cap
  geom_rect(
    data = ink_plot,
    aes(
      xmin = col - 0.21,
      xmax = col + 0.21,
      ymin = row + 0.29,
      ymax = row + 0.41
    ),
    fill = "#252321",
    color = "#151413",
    linewidth = 0.35
  ) +
  
  # cap highlight
  geom_rect(
    data = ink_plot,
    aes(
      xmin = col - 0.17,
      xmax = col - 0.11,
      ymin = row + 0.30,
      ymax = row + 0.40
    ),
    fill = alpha("white", 0.16),
    color = NA
  ) +
  
  # glass highlight
  geom_rect(
    data = ink_plot,
    aes(
      xmin = col - 0.31,
      xmax = col - 0.22,
      ymin = row - 0.32,
      ymax = row + 0.03
    ),
    fill = alpha("white", 0.14),
    color = NA
  ) +
  
  # paper bottle label
  geom_rect(
    data = ink_plot,
    aes(
      xmin = col - 0.29,
      xmax = col + 0.29,
      ymin = row - 0.20,
      ymax = row - 0.01
    ),
    fill = alpha("#FCFAF5", 0.94),
    color = "#D8D0C4",
    linewidth = 0.25
  ) +
  
  # Japanese ink name
  geom_text(
    data = ink_plot,
    aes(x = col, y = row - 0.08, label = ink_name_japanese),
    family = "Hiragino Sans",
    fontface = "bold",
    size = 4.8,
    color = "#26231F"
  ) +
  
  # romaji name
  geom_text(
    data = ink_plot,
    aes(x = col, y = row - 0.16, label = ink_name),
    family = "Baskerville",
    size = 2.7,
    color = "#4A443E"
  ) +
  
  # shop sign background
  annotate(
    "rect",
    xmin = 1.72,
    xmax = 3.28,
    ymin = 3.50,
    ymax = 3.66,
    fill = "#F8F3EA",
    color = NA
  ) +
  
  # Japanese shop name
  annotate(
    "text",
    x = 2.12,
    y = 3.58,
    label = "色彩雫",
    family = "Hiragino Sans",
    fontface = "bold",
    size = 5.4,
    color = "#4F443A"
  ) +
  
  # English shop name
  annotate(
    "text",
    x = 2.90,
    y = 3.58,
    label = "INK SHOP",
    family = "Baskerville",
    size = 5.4,
    color = "#4F443A"
  ) +
  
  scale_fill_identity() +
  
  coord_equal(
    xlim = c(-0.95, 5.95),
    ylim = c(-0.82, 3.88),
    clip = "off"
  ) +
  
  theme_void() +
  
  theme(
    plot.background = element_rect(fill = "#F7F2E9", color = NA),
    panel.background = element_rect(fill = "#F7F2E9", color = NA),
    plot.margin = margin(15, 20, 15, 20)
  )
```

## The palette as data

And because this is still a data project, here is the resulting palette in the same perceptual order.

```{r}
#| label: ink-table
#| echo: false
#| warning: false
#| message: false

library(kableExtra)

ink_table <- ink_plot |>
  transmute(
    `日本語` = ink_name_japanese,
    Ink = ink_name,
    Meaning = description,

    Hex = cell_spec(
      hex,
      format = "html",
      background = hex,
      color = if_else(L > 65, "#222222", "#FFFFFF"),
      bold = TRUE
    ),

    `Hue °` = round(H),
    Chroma = round(C, 1),
    Lightness = round(L, 1)
  )

ink_table |>
  knitr::kable(
    format = "html",       # <-- IMPORTANT
    escape = FALSE,
    align = c(
      "c", "l", "l", 
      "r", "r", "r", "c"
    )
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c(
      "striped",
      "hover",
      "condensed",
      "responsive"
    ),
    full_width = TRUE
  ) |>
  kableExtra::column_spec(
    1,
    bold = TRUE,
    width = "6em"
  ) |>
  kableExtra::column_spec(
    2,
    width = "8em"
  ) |>
  kableExtra::column_spec(
    3,
    width = "15em"
  ) |>
  kableExtra::column_spec(
    4,
    width = "5em"
  )
```

I started with an old R script containing 24 fountain pen colours.

Somewhere along the way I learned a little more about perceptual colour spaces and built a tiny imaginary Japanese ink shop out of geom_rect().

Now I just want more ink.
🖋️
