---
title: "Creating a Seigaha Motif with ggplot2"
date: "2024-12-21"
categories: [r, ggplot]
format:
  html:
    code-fold: true
execute:
  echo: true
---

## Seigaiha 青海波 🌊

The Seigaiha (青海波) motif, literally translated as "blue ocean waves" is a classic Japanese pattern used in textiles, ceramics, kimono and art. It features stylized waves  created with concentric circles forming arches. I just learned that it symbolizes surges of good luck - delightful and thought great start to have it for blog post. 

-   青 = Blue
-   海 = Ocean
-   波 = Wave

Using `ggplot2` and additional packages like `ggforce` and c`owplot,` I was able to recreate this timeless design programmatically.  

### Walk Through of How I Created the Pattern

First need to load up the required libraries.  Recently, I discovered package called [`annotater`](https://annotater.liomys.mx/) , which let you add annotation to the packages you use in script! It's such a useful tool - future me (and anyone collaborating on my code) will likely thank me for using it.  


```{r}

library(tidyverse) # Easily Install and Load the 'Tidyverse' 
library(ggforce) # Accelerating 'ggplot2'   
library(cowplot) # Streamlined Plot Theme and Plot Annotations for 'ggplot2'  
library(gt) # Easily Create Presentation-Ready Display Tables 

col10 <- str_split("012a4a-013a63-01497c-014f86-2a6f97-2c7da0-468faf-61a5c2-89c2d9-a9d6e5","-")

col10_pal <-str_c("#",col10 |> unlist())
```

### Generating the Grid Data

We create a grid of coordinates to serve as the base for our motif.  To archieve the characteristic offset of Seigaiha waves, we adjust the `x` coordinates for odd rows: 


```{r}
df <- expand_grid(x = seq(-16, 16, by = 2), 
                  y = seq(-10, 10, by = 1)) |>
  arrange(y,x) |>
  mutate(y_odd=(y%%2==1)) |>
  mutate(idx=row_number()-1) |>
  mutate(x=if_else(y_odd,x+1,x))

```

### Expanding for Circle Sizes
The radii of the circles vary in five steps.  It could be smaller or larger. I could've also add randomness to it too, but I just kept it constant for now. 


```{r}
r_values <- seq(0.3,1.1,length.out=5)
df_long <- df |> expand_grid(r=r_values)


df_long <- df_long |>
  mutate(r_var=if_else(idx%%5==0,sqrt(r),r))

#range(df_long$x)
#range(df_long$y)

df_long |> head() |> gt()
```

### Plotting the Seigaiha Motif

Using `geom_circle` from `ggforce`, we layer circles over the grid.  We map the fill colour to the adjusted radius (r_var) for a slight gradeient effect:  

```{r}
df_long |>
  arrange(-y, idx, desc(r)) |>
  ggplot() +
  geom_circle(
    aes(
      x0 = x,
      y0 = y,
      r = r,
      fill = r_var ## just wanted to give bit of variance
    ),
    linewidth = 0.1,
    color = "#fffff3de",
    linetype = 3
  ) +
  coord_fixed(clip = "on",
              xlim = c(-15, 14.5),
              ylim = c(-9, 9)) +
  theme_nothing() +
  scale_fill_gradientn(colors = col10_pal) +
  scale_color_gradientn(colors = col10_pal)

```

### Final Touch 

The `theme_nothing()` from `cowplot` removes all unnecessary visual elements (axes, labels etc.), leaving only the motif.  I've adjusted `xlim` and `ylim` so that I'm clipping out the extra circles. 



